---
title: "An overview of the data and some intuition wich possibly is way off"
author: "Christian Oppegård Moen"
date: "dd-mm-yyyy"
output: 
  bookdown::pdf_document2:
    toc_depth: '3' # Depth of contents
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: TMA4285 - Time Series
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption} # add caption to figures.
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F, # Strips whitespace
  prompt=F, #?
  cache=T, # Cache for quick knit.
  # root.dir = "./subfolder",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```

```{r config, include=F}
if (F){
  # setwd("C:\Users\kikka\AppData\Local\R\win-library\4.2\rmarkdown\rmarkdown\templates\project2023_christian\skeleton") # Location for the skeleton. Might wanna change this
  options(error=recover)
  options(error=NULL)
  par(mar=c(5,4,4,2)+0.1) # default margins
}
defaultMargins = par(mar=c(5,4,4,2)+0.1) # default margins
```

```{r packages, include=F}
library(readr)
library(ggplot2)
library(nlme) # To get season slopes
library(lubridate)
library(Rmisc)
library(moments) # Skewness test
library(astsa)
```

# Task
Model inflation rate along with three covariates of our choosing.

# Exploratory Data Analysis (EDA)
Import data and format date.
```{r data, options}
pd <- read_csv("projectdataFormated.csv")
typeof(pd)
pd$Month = as.Date(paste0(sub("M", "-", pd$Month), "-01"), format = "%Y-%m-%d")
head(pd)
```

## Inflation overview
Inflation over time:

```{r inflationSeason, fig.cap="Inflation over time colored according to the four seasons. The colouerd lines are 'loess' fitted curves per season."}
pd$MonthInt = as.integer(format(pd$Month, "%m"))
pd$Season <- ifelse(month(pd$Month) %in% c(12, 1, 2), "Winter",
                   ifelse(month(pd$Month) %in% c(3, 4, 5), "Spring",
                   ifelse(month(pd$Month) %in% c(6, 7, 8), "Summer", "Autumn")))

# Season colored plot with 'loess' fitted curves ----
ggInflation = ggplot(pd, aes(x = Month, y = Inflation))+
  geom_point(aes(col = Season)) +
  geom_line(linetype = 1, lwd =0.5) +
  geom_smooth(aes(col=Season), se=F)+
  theme_minimal()
ggInflation
```
In Figure \@ref(fig:inflationSeason) the data seem normal, with timely changes in variation.
There is not much evidence of any trends related to the four seasons. It seems like summer follow spring with a slight delay, suggesting that it could be beneficial to combine these two seasons.

```{r inflationHist, fig.cap="Histogram of inflation along with a normal curve of the sample mean and variance."}
mu <- mean(pd$Inflation)
sigma <- sd(pd$Inflation)
hist(pd$Inflation, breaks = 20, prob = TRUE, xlab = "Values", ylab = "Density")
curve(dnorm(x, mean = mu, sd = sigma), col = "cyan", lwd = 1, add = TRUE)
```
The histogram in Figure \@ref(fig:inflationHist) also suggest normality. It also shows evidence of positive skewedness. The skewedness is significant per D'Agostino test. (Dunno much about this test. Guess we wont use it anyways.)

```{r agostino, options}
agostino.test(pd$Inflation)
```

```{r inflationPerSeasonTrend, fig.cap="Linear fit to each season."}
t = rep(1,10)
for (i in 2:15){
  t = c(t, rep(i, 12))
}
t = c(t, 16)
pd$SeasonCount = paste0(pd$Season, t)

ggInflationSeason = ggplot(pd, aes(x = Month, y = Inflation, col=Season))+
  geom_point() + 
  stat_smooth(aes(fill=SeasonCount), method = "lm", se=F, show.legend=F, lwd=0.5)+
  theme_minimal()
ggInflationSeason
```
It is somewhat difficult to see any trends per season in Figure \@ref(fig:inflationPerSeasonTrend), so we compute some values below and show the slopes in a Boxplot.

```{r inflationTrends, options}
# Find season slopes ----
seasonSlopes = coef(lmList(Inflation~Month|SeasonCount, data = pd))["Month"]
seasons = unique(pd$Season)

seasonTrends = data.frame(row.names = c("mean", "var"))
seasonSlopes2 = list()
for (s in seasons){
  st = na.omit(seasonSlopes$Month[startsWith(rownames(seasonSlopes), s)])
  seasonSlopes2[[s]] = st
  seasonTrends[[s]] = c(mean(st), var(st))
}
seasonTrends
```

```{r inflationBox, fig.cap="Boxplot of the slopes of inflation per season. Confidence interval and mean for Autumn is shown by dashed grey lines."}
boxplot(seasonSlopes2, notch = T)
abline(h=0, lty=2)
# abline(h=seasonTrends['mean', "Winter"])
abline(h=CI(seasonSlopes2$Autumn, ci=0.95), lty = 2, col="gray")
```

Figure \@ref(fig:inflationBox) suggest that Autumn is significantly negative. That is, there is a significant decrease in inflation during the Autumn. Notches say that it is significantly different from the other seasons. Why negative inflation in the autumn tho? Saving money for the holidays?



## Inflation Time Series

Let's look at acf, pacf and so on.

```{r temp, fig.cap="ACF and PACF of inflation compared to a samesized sample of normal data."}
set.seed(420)
par(mfrow=c(2,2))
normalData = rnorm(dim(pd)[1])
acf(pd$Inflation, lag.max = 25)
pacf(pd$Inflation, lag.max = 25)
acf(normalData, lag.max = 25)
pacf(normalData, lag.max = 25)
```
ACF show somewhat cut off immediately, but we notice the significant 4, 12 and 24 lag.

PACF show that the 4, 8, 12, 13 and 24 lag is significant.

# Covariate/Model selection
Not much to go on but the normality shown in figure \@ref(fig:inflationHist). Suppose the inflation can be model as a regression with autocorrelated errors. Namely,
\begin{equation}
  Y_t = \sum_{i=1}^r \beta_i z_{ti} + x_t, 
\end{equation}
where $x_t$ has some covariance structure $\gamma_x(s,t)$, e.g., ARMA, $z_{ti}$ is (unknown?) covariate $i$ at time $t$ and $\beta_i$ are cofficients. In OLS we assume that the noise is Gaussian, that is, $\gamma_x(s,t) = \sigma_x^2$ for $s=t$, else zero. That is, the temporal $x_t$ captures the stationary process, but not the variance variation we discussed earlier(?). More on this in Shumway & Stoffer ch. 3.8 p.143.

Assuming that the covariates are known, we can standardize the variance by multiplying by the root of the covariance matrix. Then we get the standardized (not centralized) equation
\begin{equation}
  \boldsymbol{y}^* = \mathbf{Z}^* \boldsymbol{\beta} + \boldsymbol{\delta},
\end{equation}
where $\boldsymbol{y}^*  = \mathbf{\Gamma}^{-1/2} \boldsymbol{y}$, $\mathbf{Z}^* = \mathbf{\Gamma}^{-1/2} \mathbf{Z}$, blablabla...
Since we are in OLS territory we want to write it in matrix notation to make coefficient estimation neat and familiar by the hat-matrix. Let's try by following r code on page 144. What sarima tho, perhaps $(1,0,0)\times (P=?, D=?, Q=?)_{12}$

\textcolor{red}{Needs more work!}
```{r Sarima, options}
summary(fit <- lm(Inflation~Unemployed + Bankrupt + Consumption, data = pd)) # dayum, inline variable assignment
acf2(resid(fit), 24)
# sarima: Seasonal components: P, D, Q, S = SAR order, seasonal diff., SMA order
sarima(pd$Inflation, 1,0,0, S=12, xreg = cbind(Unemp = pd$Unemployed, Bankr = pd$Bankrupt, Cons = pd$Consumption))
```


# Future work
Tenke at Garch can vær iteressant da æ synes at d ikke e ufornuftig med litt varians analyse basert på plot \@ref(inflationSeason)
