---
title: "Title"
author: "Christian Oppegård Moen"
date: "dd-mm-yyyy"
output: 
  bookdown::pdf_document2:
    toc_depth: '3' # Depth of contents
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: code - courseName
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption} # add caption to figures.
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F, # Strips whitespace
  prompt=F, #?
  cache=T, # Cache for quick knit.
  # root.dir = "./subfolder",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```

```{r config, include=F}
if (F){
  options(error=recover)
  options(error=NULL)
  par(mar=c(5,4,4,2)+0.1) # default margins
}
defaultMargins = par(mar=c(5,4,4,2)+0.1) # default margins
```

\newpage

# GARCH(p,q) 
Definitions for notational purposes:
\begin{itemize}
  \item $\mathcal{R}_1=(r_1,...,r_{max(p,q)})$
  \item $\mathcal{R}_2=(r_t,...,r_{t+max(p,q)})$
  \item $\boldsymbol \alpha=(\alpha_0,...,\alpha_p)$ 
  \item $\boldsymbol \beta = (\beta_1,...,\beta_q)$
  \item $\boldsymbol r_{tp} = (1, r_{t-1}, ..., r_{t-p})^\top$ 
  \item $\boldsymbol \sigma_{tq}^2 = (\sigma_{t-1}^2,...,\sigma_{t-q}^p)^\top$
\end{itemize}

GARCH($p$, $q$) is given by
\begin{align}
     r_t &= \sigma_t \epsilon_t, \label{eq:garchReturn}\\
     \sigma^2_t &= \alpha_0 + \sum_{j=1}^p \alpha_j r_{t-j}^2 + \sum_{j=1}^q \beta_j \sigma_{t-j}^2. \label{eq:garchVar}\\
     \sigma_t^2 &= \boldsymbol \alpha \boldsymbol r_{tp}^2 + \boldsymbol \beta \boldsymbol \sigma_{tq}^2.
\end{align}

We know that $r_t|\mathcal{R}_2\sim \mathcal{N}(0, \sigma_t^2)$

The log likelihood $l( \boldsymbol{\alpha}, \boldsymbol{\beta} | \mathcal{R}_1 ) \propto -\ln L(\boldsymbol{\alpha}, \boldsymbol{\beta} | \mathcal{R}_1 )$ is given by
\begin{equation}
  l( \boldsymbol{\alpha}, \boldsymbol{\beta} | \mathcal{R}_1 ) = \sum_{t=m+1}^n \ln(\sigma_t^2)) + \frac{r_t^2}{\sigma_t^2} 
\end{equation}

In the following chunk we implement the so called critical function.

```{r critFunc, options}
garchLL = function(params, r, p=1, q=0){
  
  n = length(r)
  m = max(p,q)
  
  alpha = exp(params[1:(p+1)])
  if (q==0){beta = 0}
  else {beta = exp(params[-(1:(p+1))]) }
  
  # initialize variance and set first m values
  sigma_sq = numeric(n)
  sigma_sq[1:m] = t(alpha) %*% c(1,r[p:1]^2) # should they be zero?
  # sigma_sq[1:m] = 0 # says so above eq. (5.52)
  
  # Iteratively compute each variance
  for (t in (m+1):n){
    sigma_sq[t] = sum(alpha * c(1,r[(t-1):(t-p)]^2)) + sum(beta*sigma_sq[(t-1):(t-q)]) 
  }
  ll = sum(log(sigma_sq) + r^2/sigma_sq)
  return(ll)
}
```

The variance $\sigma_t^2$ can be estimated by \textit{one-step-ahead} forecasting given by
\begin{equation}
  \hat\sigma_{t}^2 = \hat{\boldsymbol \alpha} \boldsymbol r_{tp}^2 + \hat{\boldsymbol \beta} \hat{\boldsymbol \sigma}_{tq}^2
\end{equation}

```{r sigmaForecast, options}
sigmaForecast = function(mod){
  n = mod$n
  m = mod$m
  p = mod$p
  r = mod$r
  
  alpha = mod$estim[1:(p+1)]
  if (mod$q==0){beta = 0; q=1} # If ARCH(p) model
  else{beta = mod$estim[-(1:(p+1))]; q = mod$q}
  
  sf = numeric(n) # sigma squared forecasts
  # sf[1:p] = sum(alpha*c(1,r[p:1]^2))
  # sf[1:q] = 0
  # sf[1:m] = alpha[1]/(1-sum(c(alpha, beta))) # GPT
  sf[1:m] = sum(alpha*c(1,r[1:p]^2)) # used in sim test
  
  for (t in (m+1):n){
    sf[t] = sum(alpha * c(1,r[(t-1):(t-p)]^2)) + sum(beta*sf[(t-1):(t-q)])
    if(is.na(sf[t])){browser()}
  }
  return(sf)
}

# TODO: make garch function
garch = function(r, p=1, q=0, init=c(0.1,0.1)){
  n = length(r)
  
  alpha = init[1:(p+1)]
  if (q==0){beta = 0; q=1} # If ARCH(p) model
  else{beta = init[-(1:(p+1))]; q = q}
  
  estim = exp(optim(par = log(init), fn = garchLL, r = r, p=p, q=q, method = "BFGS")$par)
  
  mod = list(
    p = p,
    q = q,
    n = n,
    m = max(p,q),
    init = init,
    estim = estim,
    r=r
  )
  
  mod$sigmaForecast = sigmaForecast(mod)
  return(mod)
  
}

modResults = function(mod, main="", plot=T){
  # print(rbind(estim = mod$estim, init = mod$init))
  if (plot){
    plot(mod$r, type="l", main=main)
    lines(mod$sigmaForecast, col="cyan")
  }
}
```



# Tests
```{r testFunction, options}
testModel = function(mod){
  n = mod$n
  m = mod$m
  p = mod$p
  
  alpha = mod$paramSim[1:(p+1)]
  if (mod$q==0){beta = 0; q=1}
  else{beta = mod$paramSim[-(1:(p+1))]; q = mod$q}
  
  r = numeric(n)
  r[1:m] = rnorm(m)
  sigma_sq = numeric(n)
  sigma_sq[1:m] = sum(alpha*c(1,r[1:p]^2))
  # sigma_sq[1:m] = alpha[1]/(1-sum(c(alpha, beta))) # GPT
  
  # browser()
  for (t in (m + 1):n){
    sigma_sq[t] = sum(alpha*c(1,r[(t-1):(t-p)]^2)) + sum(beta*sigma_sq[(t-1):(t-q)])
    r[t] = rnorm(1, sd=sqrt(sigma_sq[t]))
  }
  # browser()
  # estimation
  estim = exp(optim(par = log(mod$init), fn = garchLL, r = r, p=p, q=q, method = "BFGS")$par)
  comparison = (rbind(real = mod$paramSim, estim = estim, init = mod$init))
  return(list(r = r, estim = estim, comparison = comparison))
}


```

```{r arch1, options}
set.seed(420)
arch1 = list(
  p = 1, 
  q= 0,
  m= 1,
  n = 1e4,
  init = rep(0.1,2),
  paramSim = c(0.01, 0.2)
)


testResult = testModel(arch1)
testResult$comparison

# qqnorm(testResult$r)
# plot(testResult$r)
```

```{r garch12, options}
set.seed(420)
garch12 = list(
  p = 1, 
  q= 2,
  m= 2,
  n = 1e4,
  init = rep(0.1,4),
  paramSim = c(0.01, 0.2, 0.1, 0.5)
)


testResult = testModel(garch12)
garch12$estim = testResult$estim
round(testResult$comparison,3)

garch12$r = testResult$r
garch12$sigmaForecast = sigmaForecast(garch12)
plot(garch12$r, type = "l")
lines(garch12$sigmaForecast, col="cyan")
```

# Real data (HOW YOU FIT IT AND VIEW IT)
```{r loadData, options}
df <- read.csv('projectdata.csv', header = T, sep=";", dec=",", stringsAsFactors=FALSE)


# qqnorm(r)
# plot(r, type = "l")
```

```{r garch, fig.height=7}
# make data to fit
fit = lm(Inflation~Unemployed+Consumption+InterestRate, data = df)
r = df$Inflation - fit$fitted.values
d = diff(df$Inflation)
par(mfrow=c(2,1))
# inflation - regression
garch12regr = garch(r, p=1, q=2, init=rep(0.1, 1+2+1))
modResults(garch12regr, main="garch(1,2) on regression")

# diff inflation
garch12 = garch(d, p=1, q=2, init=rep(0.1, 1+2+1))
modResults(garch12, "garch(1,2) on diff")
par(mfrow=c(1,1))
```




```{r arch, fig.height=9}
par(mfrow=c(3,2))
arch1regr = garch(r, p=1, q=0, init=rep(0.1, 2))
modResults(arch1regr, main="arch(1) on inflation - regression")

arch1d = garch(d, p=1, q=0, init=rep(0.1, 2))
modResults(arch1d, main = "arch(1) on diff(inflation)")

arch2regr = garch(r, p=2, q=0, init=rep(0.1, 3))
modResults(arch2regr, main="arch(2) on inflation - regression")

arch2d = garch(d, p=2, q=0, init=rep(0.1, 3))
modResults(arch2d, "arch(2) on diff(inflation)")

arch10regr = garch(r, p=10, q=0, init=rep(0.1, 11))
modResults(arch10regr, main="arch(10) on inflation - regression")

arch10d = garch(d, p=10, q=0, init=rep(0.1, 11))
modResults(arch10d, "arch(10) on diff(inflation)")
par(mfrow=c(1,1))

list(
arch1regr = arch1regr$estim,
arch1d = arch1d$estim,
arch2regr = arch2regr$estim,
arch2d = arch2d$estim,
arch10regr = arch10regr$estim,
arch10d = arch10d$estim
)

```







