---
title: "Title"
author: "Christian Oppegård Moen"
date: "dd-mm-yyyy"
output: 
  bookdown::pdf_document2:
    toc_depth: '3' # Depth of contents
    number_sections: false
  # pdf_document:
  # #   toc: no
  #   toc_depth: '3'
subtitle: code - courseName
urlcolor: blue
editor_options: 
  chunk_output_type: console
header-includes:
- \usepackage[width=0.8\textwidth]{caption} # add caption to figures.
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = T, tidy=T, message=F, warning=F,
  strip.white=F, # Strips whitespace
  prompt=F, #?
  cache=T, # Cache for quick knit.
  # root.dir = "./subfolder",
  size="scriptsize",
  fig.width=7, 
  fig.height=5, 
  fig.align = "center"
)
```

```{r config, include=F}
if (F){
  options(error=recover)
  options(error=NULL)
  par(mar=c(5,4,4,2)+0.1) # default margins
}
defaultMargins = par(mar=c(5,4,4,2)+0.1) # default margins
```


\newpage
```{r libraries, options}
source("chrisFunctions.r")
library(ggplot2)
```

# GARCH
GARCH($p$, $q$) is given by
\begin{align}
     r_t &= \sigma_t \epsilon_t, \label{eq:garchReturn}\\
     \sigma^2_t &= \alpha_0 + \sum_{j=1}^p \alpha_j r_{t-j}^2 + \sum_{j=1}^q \beta_j \sigma_{t-j}^2. \label{eq:garchVar}
\end{align}

It can be shown that a GARCH($p$, $q$) admits a non-Gaussian ARMA($p$, $q$) for the squared process
\begin{equation}
  r_t^2
\end{equation}



```{r loadData, include=F}
df <- read.csv('projectdata.csv', header = T, sep=";", dec=",", stringsAsFactors=FALSE)
# head(df)
```

# acf and pacf
For fun we consider the simple acf and pacf of the difference
\begin{equation*}
  \nabla x_t = x_t - x_{t-1}
\end{equation*}
```{r simpleDiff, fig.cap="Manual to the left and built-in to the right"}
d = diff(df$Inflation)
par(mfrow=c(2,2))
dacf = sampleAcf(d,22)
dacf2 = acf(d)

dpacf = samplePacf(d, 22)
dpacf2 = pacf(d)
```

Now let's look at the return
\begin{equation}\label{eq:squaredDiff}
  \nabla r_t = \frac{x_t - x_{t-1}}{x_{t-1}}.
\end{equation}
Problem: Some values are zero. Let's make sure that they are not. (Should not matter since we are looking at variance, right?)
```{r return, fig.cap = "Return"}
n = dim(df)[1]
shift = min(df$Inflation - 0.1)
inflationShifted = df$Inflation - shift
r = diff(inflationShifted)/inflationShifted[1:(n-1)]
par(mfrow=c(1,2))
dacf = sampleAcf(r,22)

dpacf = samplePacf(r, 22)
```

Finally, let's consider the squared return, which is supposed to be ARMA(p,q)
```{r sqReturn, fig.cap="Squared return"}
par(mfrow=c(1,2))
dacf = sampleAcf(r^2,22)

dpacf = samplePacf(r^2, 22) # Verified with pacf fnc.
```

\begin{equation}
\begin{array}{l}
\text { Table 3.1. Behavior of the ACF and PACF for ARMA models }\\
\begin{array}{cccc}
\hline \hline & \operatorname{AR}(p) & \operatorname{MA}(q) & \operatorname{ARMA}(p, q) \\
\hline \text { ACF } & \text { Tails off } & \begin{array}{c}
\text { Cuts off } \\
\text { after lag } q
\end{array} & \text { Tails off } \\
\text { PACF } & \begin{array}{c}
\text { Cuts off } \\
\text { after lag } p
\end{array} & \text { Tails off } & \text { Tails off } \\
\hline
\end{array}
\end{array}
\end{equation}



<!-- ```{r gptAns, options} -->
<!-- # Function to fit GARCH(p, q) model without using non-base libraries -->
<!-- fit_garch <- function(data, p, q, omega_init, alpha_init, beta_init) { -->
<!--   # Length of the data -->
<!--   n <- length(data) -->

<!--   # Initialize parameters -->
<!--   omega <- omega_init -->
<!--   alpha <- alpha_init -->
<!--   beta <- beta_init -->

<!--   # Initialize arrays to store the squared returns and conditional variances -->
<!--   h <- numeric(n) -->
<!--   epsilon_squared <- numeric(n) -->

<!--   # Calculate squared returns and initialize conditional variances -->
<!--   epsilon_squared[1:(p + q)] <- data[1:(p + q)]^2 -->
<!--   h[1:(p + q)] <- mean(epsilon_squared[1:(p + q)]) -->

<!--   # Main loop to estimate parameters and conditional variances -->
<!--   for (t in (p + q + 1):n) { -->
<!--     # Estimate conditional variance using GARCH(p, q) model -->
<!--     h[t] <- omega + sum(alpha * epsilon_squared[(t - p):(t - 1)] + beta * h[(t - q):(t - 1)]) -->

<!--     # Calculate the squared return at time t -->
<!--     epsilon_squared[t] <- data[t]^2 -->
<!--   } -->

<!--   # Return the estimated parameters and conditional variances -->
<!--   result <- list(omega = omega, alpha = alpha, beta = beta, conditional_variances = h) -->
<!--   return(result) -->
<!-- } -->

<!-- # Example usage: -->
<!-- # Replace 'your_data' with your actual time series data -->
<!-- data <- df$inflation -->
<!-- # Replace p, q, omega_init, alpha_init, beta_init with your desired values -->
<!-- result <- fit_garch(data, p = 1, q = 1, omega_init = 0.01, alpha_init = 0.2, beta_init = 0.3) -->

<!-- ``` -->

# GARCH(p,q) Implementation
Definitions for notational purposes:
\begin{itemize}
  \item $\mathcal{R}_1=(r_1,...,r_{max(p,q)})$
  \item $\mathcal{R}_2=(r_t,...,r_{t+max(p,q)})$
  \item $\boldsymbol \alpha=(\alpha_0,...,\alpha_p)$ 
  \item $\boldsymbol \beta = (\beta_1,...,\beta_q)$
  \item $\boldsymbol r_{tp} = (1, r_{t-1}, ..., r_{t-p})^\top$ 
  \item $\sigma_{tq}^2 = (\sigma_{t-1}^2,...,\sigma_{t-q}^p)^\top$
\end{itemize}

It can be shown that $r_t|\mathcal{R}_2\sim \mathcal{N}(0, \sigma_t^2)$, $\mathcal{R}_2=(r_t,...,r_{t+max(p,q)})$. Because of the normality of the conditional return we can find MLE of $\boldsymbol \alpha$ and $\boldsymbol \beta$ parameters. That is, the likelihood to be maximized is given by


 \begin{align}
     L(\boldsymbol \alpha, \boldsymbol \beta | \mathcal{R}_1) &= \prod_{t = \max(p,q)+1}^n f_{\boldsymbol \alpha, \boldsymbol \beta}(r_t|\mathcal{R}_2) \nonumber \\
     &=\prod_{t = \max(p,q)+1}^n (2\pi \sigma_t^2)^{-1/2} \exp(-\frac{1}{2} \frac{r_t^2}{\sigma_t^2}),
 \end{align}
 
 where $\mathcal{R}_1=(r_1,...,r_{max(p,q)})$ and $\sigma_t^2$ is given by equation \eqref{eq:garchVar}. We define the criterion function $l$ to be minimized as proportional to $-\ln L$, such that
 
 \begin{align}
     l(\boldsymbol \alpha, \boldsymbol \beta|\mathcal{R}_1) &= \sum_{t = \max(p,q)+1}^n \ln(\sigma_t^2) + \frac{r_t^2}{\sigma_t^2} \nonumber \\
     &= \sum_{t = \max(p,q)+1}^n \ln (\alpha_0 + \sum_{j=1}^p \alpha_j r_{t-j}^2 + \sum_{j=1}^q \beta_j \sigma_{t-j}^2) + \frac{r_t^2}{\alpha_0 + \sum_{j=1}^p \alpha_j r_{t-j}^2 + \sum_{j=1}^q \beta_j \sigma_{t-j}^2} \nonumber \\
     &=  \sum_{t = \max(p,q)+1}^n \ln(\boldsymbol \alpha \boldsymbol r_{tp} + \boldsymbol \beta \boldsymbol \sigma_{tq}^2) + \frac{r_t^2}{\boldsymbol \alpha \boldsymbol r_{tp} + \boldsymbol \beta \boldsymbol \sigma_{tq}^2},
 \end{align}
 
 where $\boldsymbol r_{tp} = (1, r_{t-1}, ..., r_{t-p})^\top$ and $\sigma_{tq}^2 = (\sigma_{t-1}^2,...,\sigma_{t-q}^p)^\top$



```{r garch(pq), options}
critFunc = function(params, r, p=1, q=1){
  # LogLike of alpha and beta given the first r_t values t=(1,..., max(p,q))(Eq. 5.46)
  # Defaults to GARCH(1,1) and uses only the first three params (alpha0, alpha1, beta1)
  # params: vector of parameters. First p+1 are alphas, rest are betas
  # r: vector of returns (Eq. 5.34)
  # p,q: GARCH(p,q) model parameters
  
  n = length(r)
  m = max(p,q)
  
  # If else statements in case it is not "generalized" (ARCH(p))
  if (p == 0){
    # Prolly not needed, never gonna remove the AR part?
    alpha = c(0)
    p=1
  }
  else {alpha = params[1:(p+1)]}
  if (q == 0){
    beta = c(0)
    q = 1
  }
  else {beta = params[(p+2):(1+p+q)]}
  
  # Initialize first m conditional variances.
  # TODO: This is not correct initialization of the first sigma[1:m], but it works for now.
  sigma = numeric(n)
  sigma[1:m] = (r[1:m]^2)
  
  ll = 0 # log likelihood (objective is to minimize this)
  for (t in (m+1):n){
    sigma[t] = t(alpha) %*% c(1,r[(t-1):(t-p)]^2) + t(beta) %*% sigma[(t-1):(t-q)]
    ll = ll + log(sigma[t]) + r[t]^2/(sigma[t])
  }
  return(ll)
}

# r shifted ----
# Shifting inflation since r (eq. 5.34) cannot have zeroes in the denominator.
set.seed(420) # Why? don't ask me
shift = min(df$Inflation - 0.1) # inflation seems to be discretized per 0.1
inflationShifted = df$Inflation - shift
r = diff(inflationShifted)/inflationShifted[1:(n-1)]

# Test garch(1,2) ----
alpha_init = c(0.1,0.1)
beta_init = c(0.1,0.1)
p = length(alpha_init)-1
q = length(beta_init)
loglike = critFunc(c(alpha_init, beta_init), r, p=p, q=q)
opt = optim(par=c(alpha_init, beta_init), fn=critFunc, r=r, p=p, q=q)
alpha = opt$par[1:(p+1)]
beta = opt$par[(p+2):(1+p+q)]
cbind(alpha, beta)

# Sigma estimation ----
sigmaForecast = function(alpha, beta, r){
  # One-step-ahead forecasts of the volatility sigma (Eq. 5.52)
  # alpha = (alpha_0,..., alpha_p): Estimated parameters by num. optim.
  # beta = (beta_1,..., beta_q): Estimated parameters by num. optim.
  # r: vector of returns.
  
  p = length(alpha) - 1
  q = length(beta)
  m = max(p,q)
  
  # Initialize first m conditional variances.
  # TODO: This is not correct initialization of the first sigma[1:m], but it works for now.
  sigma = numeric(n)
  sigma[1:m] = (r[1:m]^2)
  for (t in (m+1):n){
    sigma[t] = t(alpha) %*% c(1,r[(t-1):(t-p)]^2) + t(beta) %*% sigma[(t-1):(t-q)]
  }
  return(sigma)
}

sigma_estim = sigmaForecast(alpha, beta, r)
plot(r, type = 'l', )
lines(sigma_estim[2:length(r)], col="cyan", lty=2)
plot(df$Inflation, type = "l")
```


```{r arch1TestSample, options}
alphas = c(0.3,0.7)
p = length(alpha)-1
n = 1e4
rs = numeric(n) # r sample
for (t in (p+1):n){
  rs[t] = rnorm(1, sd=sqrt(t(alphas)%*%c(1,rs[(t-1):(t-p)]^2)))
}
optAlph = optim(par=c(0.1,0.1), fn=critFunc, r=rs, p=p, q=0)

```

```{r gptGarch22test, options}
# Set GARCH(2,2) parameters
p <- 2
q <- 2
omega <- 0.01
alpha1 <- 0.2
alpha2 <- 0.1
beta1 <- 0.3
beta2 <- 0.2

# Set the length of the time series
n <- 1e4

# Initialize vectors to store simulated data
returns <- numeric(n)
conditional_variances <- numeric(n)

# Set initial values
returns[1:2] <- rnorm(2)
conditional_variances[1:2] <- omega / (1 - alpha1 - alpha2 - beta1 - beta2)  # Ensure that it's a valid GARCH process

# Simulate GARCH(2,2) process
for (i in 3:n) {
  epsilon <- rnorm(1)
  conditional_variances[i] <- omega + alpha1 * returns[i-1]^2 + alpha2 * returns[i-2]^2 +
                              beta1 * conditional_variances[i-1] + beta2 * conditional_variances[i-2]
  returns[i] <- epsilon * sqrt(conditional_variances[i])
}


initParams = c(0.1,0.1,0.1,0.1,0.1)
params = optim(par=c(0.1,0.1,0.1,0.1,0.1), fn=critFunc, r=returns, p=2, q=2, method = "BFGS")$par
round(rbind(real = c(omega,alpha1,alpha2,beta1,beta2), estim = params, init = initParams),2)
```
